The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-16 paper/iphas-dr2.tex
Finished polishing Sec 6

Anyone interested in the alternative detections of a source
-- hereafter called the \emph{secondary} detections --
can nevertheless obtain this information in two ways.
To begin with, whenever a secondary detection was collected 
within 10 minutes of the primary,
we have included the identifier and the photometry
of that secondary detection
in the catalogue for convenience
(columns \emph{sourceID2}, \emph{fieldID2}, 
\emph{r2}, \emph{i2}, \emph{ha2},
\emph{rErr2}, \emph{iErr2}, \emph{haErr2}, \emph{errBits2}).
Second, detections not included in the catalogue
are made available on our website.

Primary detections have been selected from all available detections using a so-called \emph{seaming} procedure,
which we adapted from the algorithm developed for the WSA\footnote{http://surveys.roe.ac.uk/wsa/dboverview.html\#merge}.
In brief, the first step is to identify all the duplicate detections
by cross-matching the overlap regions of all field catalogues,
again using a maximum matching distance of 1\arcsec.
The duplicate detections for each unique source are then ranked according to
(i) filter coverage, (ii) quality score (column \emph{errBits}),
and (iii) the average seeing of stars in the CCD frame rounded to 0.2~arcsec.
If this ranking scheme reveals multiple `winners' of seemingly identical quality,
then the one that was observed closest to the optical axis of the camera is chosen.

@@

As the final step, the primary detections
selected above were compiled
into the 98-column source catalogue
that is described in Appendix~\ref{app:columns}
and is made available on our website and through Vizier.
The original unweeded list of sources naturally included 
a significant number of spurious entries
as a result of the very sensitive default detection settings
that are employed by the CASU pipeline.
To limit the size of the source catalogue,
we have decided to enforce three basic criteria
which must be met for a candidate source
to be included in the catalogue:
\begin{enumerate}
\item the source must have been detected at S/N$>5$ in at least
one of the bands, i.e. it is required that at least one of
\emph{rErr}, \emph{iErr} or \emph{haErr} is smaller
than 0.2 mag;
\item the shape of the source must not be an obvious
cosmic ray or noise artefact, i.e. we require
either \emph{pStar} or \emph{pGalaxy} to be
greater than 20 per cent;
\item the source must not have been detected in one of the strongly
vignetted corners of the detector, 
not have had any known bad pixels in the aperture,
and not have been on the edge of one of the CCDs
(this information is contained in the \emph{errBits} quality score column, which is used for internal bookkeeping).
\end{enumerate}

\begin{table}
    \caption{Breakdown of catalogue entries as a function of the band(s)
             in which the source was detected.
             We also show the fraction of `confirmed' objects,
             which we define as those sources detected in more than one field
             (usually the field pair partner).
             We also show the average magnitudes in each class.}
    \label{tbl:detections}
    \begin{center}
        \begin{tabular}{lrcccc}
        \toprule
        Band(s) & Sources & Confirmed & $\overline{r}$ & $\overline{i}$ & $\overline{\ha}$ \\
        & [$10^6$] & [\emph{nObs}$>$1] & [mag] & [mag] & [mag] \\
        \midrule
        $r$, $i$, \ha & 159 & 91\% & 19.3 & 18.2 & 18.8  \\
        $r$, $i$ & 25 & 77\% & 21.2 & 19.9 & - \\
        $i$, \ha & 3 & 73\% & - & 19.6 & 20.2 \\
        $r$, \ha & 2 & 65\% & 20.6 & - & 20.1 \\
        $i$ & 15 & 43\% & - & 19.8 & - \\
        $r$ & 9 & 27\% & 21.1 & - & - \\
        \ha & 6 & 12\% & - & - & 19.9 \\
        \midrule
        Total  & 219 & 81\% & 19.6 & 18.6 & 18.9 \\
        \bottomrule
        \end{tabular}
        % riha: 158831809 / 144339044
		% ri: 24558323 / 18878201
		% iha: 3422905 / 2495655
		% rha: 1606867 / 1036607
		% i: 15463660 / 6612351
		% r: 9263422 / 2478774
		% ha: 5844538 / 727732
	\end{center}
\end{table}

@@

Table~\ref{tbl:detections} details the number of bands
in which the catalogue entries are captured;
159 million sources are detected in all three filters (73 per cent),
30 million are detected in two filters (14 per cent),
and the remaining 30 million are single-band detections.
Table~\ref{tbl:detections} also presents the average magnitudes
and the fraction of `confirmed' objects,
which we define as those sources which have been detected in more than
one field
(recall that the majority of the survey area
is observed twice due to the field pair strategy).
We find that the single-band detections tend to show low confirmation rates,
which suggests that a significant fraction of these entries may be spurious detections.
Hence, we strongly recommend users not to rely on single-band objects
without inspecting the image data by eye.

@@

\begin{table*}
\vspace{2cm}
\caption{Recommended minimum quality criteria 
for selecting objects with reliable colours 
from the IPHAS DR2 source catalogue. 
86~million entries in the catalogue (39\%)
satisfy all the criteria listed in this table.
For convenience, these have been flagged in the catalogue
using the column named \emph{reliable}.}
\begin{tabular}{p{8cm}lp{6cm}}
  \hline
  Quality criterion & Rows passed & Description \\
  \hline
  $rErr\,<\,0.1$ {\sc and} $iErr\,<\,0.1$ {\sc and} $haErr\,<\,0.1$ &
  109 million (50\%) &
  Require the photon noise to be less than 
  0.1 mag in all bands (i.e. S/N$>$10).
  This implicitly requires a detection in all three bands.  \\
  $r>13$ {\sc and} $i>12$ {\sc and} \ha\,$>12.5$ {\sc and not} \emph{saturated} &
  158 million (72\%) &
  The brightness must not exceed the nominal saturation limit
  and the peak pixel height must not exceed 55\,000 counts.
  Again, this implicitly requires a detection in all three bands.
  \\
  $|r- rAperMag1| 
  < 3\sqrt{rErr^2 + rAperMag1Err^2} 
  + 0.03$ &
  176 million (80\%) &
  Require the $r$ magnitude measured 
  in the default 2\farcs 3-diameter aperture
  to be consistent with the measurement 
  made in the smaller 1\farcs 2 aperture,
  albeit tolerating a 0.03 mag systematic error.
  This will reject sources for which the background
  subtraction or the deblending procedure
  was not performed reliably. \\
  $|i- iAperMag1| 
  < 3\sqrt{iErr^2 + iAperMag1Err^2} 
  + 0.03$ &
  183 million (84\%) &
  Same as above for $i$. \\
  $|ha- haAperMag1| < 
  3\sqrt{haErr^2 + haAperMag1Err^2} 
  + 0.03$ &
  158 million (72\%) &
  Same as above for \ha. \\
  \hline
  All of the above (flagged as {\bf\emph{reliable}}) &
  86 million (39\%) & \\
  \hline
\end{tabular}
\label{tab:reliable}

@@

   {\sc not} \emph{brightNeighb} &
   216 million (99\%) &
   There is no star brighter than $V < 4$ within 10 arcmin, 
   or brighter than $V < 7$ within 5 arcmin.
   Such very bright stars cause scattered light and diffraction spikes,
   which may add systematic errors to the photometry
   or even trigger spurious detections. \\  
  \hline

@@

Fig.~\ref{fig:magdist} compares the $r$-band magnitude
distributions for \emph{reliable}, \emph{veryReliable}
and unfiltered objects.
We find that 81\% of the sources 
are considered \emph{reliable}
and 54\% are \emph{veryReliable}
in the magnitude range $13 < r < 19$.
We will explain below that the
\emph{veryReliable} category is least complete
at low Galactic longitudes,
where source blending can affect
up to a quarter of the objects.
The \emph{veryReliable} flag should hence
only be used in applications in which 
completeness is not a concern.

@@

It is easy to see how the quality criteria
may be adapted to be more tolerant.
For example, by raising the allowed photometric uncertainties
from 0.1~mag to 0.2~mag,
42 million candidate sources would be added to the 109 
million satisfying the tighter error bound.
Our choice to adopt 0.1~mag as the cut-off uncertainty
for the \emph{reliable} category is a pragmatic trade-off
which we found to suit many science applications,
but certainly not all. Users of the data are encouraged 
to revise the quality criteria according to their needs.

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/uncertainties/uncertainties.pdf} 
    \caption{Mean photometric uncertainties
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Data points shown are the average values of
             columns \emph{rErr}, \emph{iErr} and \emph{haErr}
             in the catalogue 
             and the error-bars show the standard deviations.
             The dashed and solid lines indicate 
             the 10$\sigma$ and 5$\sigma$ limits, respectively.
             These statistics show the average level of the Poissonian
             photon noise and do not include systematic uncertainties.}
    \label{fig:uncertainties}
\end{figure}

Fig.~\ref{fig:uncertainties} shows the mean photometric
uncertainties (\emph{rErr}, \emph{iErr}, \emph{haErr})
as a function of magnitude.
We find the typical uncertainty to reach 0.1~mag near $r=$20.5 
and $i$/\ha$=19.5$.
We note that the fainter depth in $r$ is compensated
by the fact that most stars have brighter magnitudes in $i$ and \ha;
the average colours in the catalogue are
$\overline{(r-i)}=1.06\pm0.12$ and $\overline{(r-\ha)}=0.44\pm0.03$.
We warn that the statistics shown in Fig.~\ref{fig:uncertainties}
are the random errors based on the expected Poissonian photon noise.
Systematics, such as calibration and deblending errors,
are not included.

To appraise the level at which our photometry is affected by such systematics,
we can exploit the secondary measurements which are present in the catalogue
(i.e. \emph{r} vs \emph{r2}, \emph{i} vs \emph{i2}, \emph{ha} vs \emph{ha2}).
In Fig.~\ref{fig:pairmag} we show the mean absolute residuals between
these primary and secondary magnitudes as a function of magnitude (black dots).
We also plot the Poissonian uncertainties for comparison (solid red line).
We find the mean residual to equal $0.03\pm0.04$~mag
across the magnitude ranges 13 to 18 ($r$)
and 12 to 17 ($i$, \ha),
consistent with the accuracy of our calibration estimated above.
Stars fainter than this range appear to be dominated by photon noise (red line),
while stars at the bright end appear to suffer from large systematic
errors due to saturation effects.

>>>>

Anyone interested in the alternative detections of a source
-- hereafter called the \emph{secondary} detections --
can nevertheless obtain this information in two ways.
To begin with, whenever a secondary detection was collected 
within 10 minutes of the primary,
we have included the identifier and the photometry
of that secondary detection
in the catalogue for convenience
(columns \emph{sourceID2}, \emph{fieldID2}, 
\emph{r2}, \emph{i2}, \emph{ha2},
\emph{rErr2}, \emph{iErr2}, \emph{haErr2}, \emph{errBits2}).
Second, tables of detections not included in the catalogue
are made available on our website.

Primary detections have been selected from all available detections using a so-called \emph{seaming} procedure,
which we adapted from the algorithm developed for the WSA\footnote{http://surveys.roe.ac.uk/wsa/dboverview.html\#merge}.
In brief, the first step is to identify all the duplicate detections
by cross-matching the overlap regions of all field catalogues,
again using a maximum matching distance of 1~arcsec.
The duplicate detections for each unique source are then ranked according to
(i) filter coverage, (ii) quality score,
and (iii) the average seeing of stars in the CCD frame rounded to 0.2~arcsec.
If this ranking scheme reveals multiple `winners' of seemingly identical quality,
then the one that was observed closest to the optical axis of the camera is chosen.

@@

As the final step, the primary detections
selected above were compiled
into the final 98-column source catalogue
that is described in Appendix~\ref{app:columns}.
The original unweeded list of sources naturally included 
a significant number of spurious entries
as a result of the very sensitive default detection settings
that are employed by the CASU pipeline.
To limit the size of the source catalogue,
we have decided to enforce three basic criteria
which must be met for a candidate source
to be included in the catalogue:
\begin{enumerate}
\item the source must have been detected at S/N$>5$ in at least
one of the bands, i.e. it is required that at least one of
\emph{rErr}, \emph{iErr} or \emph{haErr} is smaller
than 0.2 mag;
\item the shape of the source must not be an obvious
cosmic ray or noise artefact, i.e. we require
either \emph{pStar} or \emph{pGalaxy} to be
greater than 20 per cent;
\item the source must not have been detected in one of the strongly
vignetted corners of the detector, 
not have had any known bad pixels in the aperture,
and not have been on the edge of one of the CCDs
(this information is contained in the \emph{errBits} quality score column, which is used for internal bookkeeping).
\end{enumerate}

\begin{table}
    \caption{Breakdown of catalogue sources as a function of the band(s)
             in which the object was detected.
             We also show the fraction of `confirmed' sources,
             which we define as those objects
             detected in more than one field
             (usually the field pair partner).
             We also show the average magnitudes for each class.}
    \label{tbl:detections}
    \begin{center}
        \begin{tabular}{lrcccc}
        \toprule
        Band(s) & Sources & Confirmed & $\overline{r}$ & $\overline{i}$ & $\overline{\ha}$ \\
        & [$10^6$] & [\emph{nObs}$>$1] & [mag] & [mag] & [mag] \\
        \midrule
        $r$, $i$, \ha & 159 & 91\% & 19.3 & 18.2 & 18.8  \\
        $r$, $i$ & 25 & 77\% & 21.2 & 19.9 & - \\
        $i$, \ha & 3 & 73\% & - & 19.6 & 20.2 \\
        $r$, \ha & 2 & 65\% & 20.6 & - & 20.1 \\
        $i$ & 15 & 43\% & - & 19.8 & - \\
        $r$ & 9 & 27\% & 21.1 & - & - \\
        \ha & 6 & 12\% & - & - & 19.9 \\
        \midrule
        Total  & 219 & 81\% & 19.6 & 18.6 & 18.9 \\
        \bottomrule
        \end{tabular}
        % riha: 158831809 / 144339044
		% ri: 24558323 / 18878201
		% iha: 3422905 / 2495655
		% rha: 1606867 / 1036607
		% i: 15463660 / 6612351
		% r: 9263422 / 2478774
		% ha: 5844538 / 727732
	\end{center}
\end{table}

@@

Table~\ref{tbl:detections} details the breakdown of these
sources as a function of the bands in which they are captured.
159 million sources are detected in all three filters (73 per cent),
30 million are detected in two filters (14 per cent),
and the remaining 30 million are single-band detections.
Table~\ref{tbl:detections} also presents the average magnitudes
and the fraction of `confirmed' objects,
which we define as those sources which have been detected in more than
one field
(recall that the majority of the survey area
is observed twice due to the field pair strategy).
We find that the single-band detections tend to show low confirmation rates,
which suggests that a significant fraction of these entries may be spurious detections.
We recommend users not to rely on single-band objects
without inspecting the image data by eye.

@@

\begin{table*}
\vspace{2cm}
\caption{Recommended minimum quality criteria 
for selecting objects with reliable colours 
from the IPHAS DR2 source catalogue. 
86~million entries in the catalogue (39\%)
satisfy all the criteria listed in this table.
For convenience, these have been flagged in the catalogue
using the column named \emph{reliable}.}
\begin{tabular}{p{8cm}lp{6cm}}
  \hline
  Quality criterion & Rows passed & Description \\
  \hline
  $rErr\,<\,0.1$ {\sc and} $iErr\,<\,0.1$ {\sc and} $haErr\,<\,0.1$ &
  109 million (50\%) &
  Require the photon noise to be less than 
  0.1 mag in all bands (i.e. S/N$>$10).
  This implicitly requires a detection in all three bands.  \\
  $r>13$ {\sc and} $i>12$ {\sc and} \ha\,$>12.5$ {\sc and not} \emph{saturated} &
  158 million (72\%) &
  The brightness must not exceed the nominal saturation limit
  and the peak pixel height must not exceed 55,000 counts.
  Again, this implicitly requires a detection in all three bands.
  \\
  $|r- rAperMag1| 
  < 3\sqrt{rErr^2 + rAperMag1Err^2} 
  + 0.03$ &
  176 million (80\%) &
  Require the $r$ magnitude measured 
  in the default 2\farcs 3-diameter aperture
  to be consistent with the measurement 
  made in the smaller 1\farcs 2 aperture,
  albeit tolerating a 0.03 mag systematic error.
  This will reject sources for which the background
  subtraction or the deblending procedure
  was not performed reliably. \\
  $|i- iAperMag1| 
  < 3\sqrt{iErr^2 + iAperMag1Err^2} 
  + 0.03$ &
  183 million (84\%) &
  Same as above for $i$. \\
  $|ha- haAperMag1| < 
  3\sqrt{haErr^2 + haAperMag1Err^2} 
  + 0.03$ &
  158 million (72\%) &
  Same as above for \ha. \\
  \hline
  All of the above (flagged as {\bf\emph{reliable}}) &
  86 million (39\%) & \\
  \hline
\end{tabular}
\label{tab:reliable}

@@

   {\sc not} \emph{brightNeighb} &
   216 million (99\%) &
   There is no star brighter than $V < 4$ within 10\arcmin, 
   or brighter than $V < 7$ within 5\arcmin.
   Such very bright stars cause scattered light and diffraction spikes,
   which may add systematic errors to the photometry
   or even trigger spurious detections. \\  
  \hline

@@

Fig.~\ref{fig:magdist} compares the $r$-band magnitude
distributions for \emph{reliable}, \emph{veryReliable}
and unfiltered objects.
We find that 81\% of the sources 
are considered \emph{reliable}
and 54\% are \emph{veryReliable}
in the magnitude range $13 < r < 19$.
We will explain below that the
\emph{veryReliable} category suffers most of
its incompleteness at low Galactic longitudes,
where source blending can affect
up to a quarter of the objects.

@@

It is easy to see how the quality criteria
may be adapted to be more tolerant.
For example, by raising the allowed photometric uncertainties
from 0.1~mag to 0.2~mag in Table~\ref{tab:reliable},
42 million candidate sources would be added to the 109 
million satisfying the tighter error bound.
Our choice to adopt 0.1~mag as the cut-off uncertainty
for the \emph{reliable} category is a pragmatic trade-off
which we found to suit many science applications,
but users are nevertheless encouraged 
to revise the quality criteria according to their needs.

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/uncertainties/uncertainties.pdf} 
    \caption{Mean photometric uncertainties
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Data points shown are the average values of
             columns \emph{rErr}, \emph{iErr} and \emph{haErr}
             in the catalogue, 
             and the error-bars show the standard deviations.
             The dashed and solid lines indicate 
             the 10$\sigma$ and 5$\sigma$ limits, respectively.
             These statistics show the average level of the Poissonian
             photon noise and do not include systematic uncertainties.}
    \label{fig:uncertainties}
\end{figure}

Fig.~\ref{fig:uncertainties} shows the mean photometric
uncertainties (\emph{rErr}, \emph{iErr}, \emph{haErr})
as a function of magnitude.
We find the typical uncertainty to reach 0.1~mag near $r=$20.5 
and $i$,\ha$=19.5$.
We note that the fainter depth in $r$ is compensated
by the fact that most stars have brighter magnitudes in $i$ and \ha;
the average colours for sources in the catalogue are
$\overline{(r-i)}=1.06\pm0.12$ and $\overline{(r-\ha)}=0.44\pm0.03$.
We warn that the statistics shown in Fig.~\ref{fig:uncertainties}
are the random errors based on the expected Poissonian photon noise.
Systematics, such as calibration and deblending errors,
are not included.

To appraise the level at which our photometry is affected by such systematics,
we can exploit the secondary measurements which are present in the catalogue
(i.e. \emph{r} vs \emph{r2}, \emph{i} vs \emph{i2}, \emph{ha} vs \emph{ha2}).
In Fig.~\ref{fig:pairmag} we show the mean absolute residuals between
these primary and secondary magnitudes as a function of magnitude (black dots).
We also plot the Poissonian uncertainties for comparison (solid red line).
We find the mean residual and standard deviation to be $0.03\pm0.04$~mag
across the magnitude ranges 13 to 18 ($r$)
and 12 to 17 ($i$, \ha).
Stars fainter than this range appear to be dominated by photon noise (red line),
while stars at the bright end appear to suffer from large systematic
errors due to saturation effects.
